### 2020-03-16 Manuscript_supershort.Rmd
SHDH comments incorporated. Title page for PNAS

@@

\‡ Corresponding author: Alejandro Damian-Serrano, email: alejandro.damianserrano@yale.edu

@@

Siphonophores, tentilla, nematocysts, predation, specialization, character evolution

@@

Predatory specialization is often associated with the evolution of modifications in the morphology of the prey capture apparatus. Specialization has been considered an evolutionary ‘dead-end’ due to the constraints associated with these morphological changes. However, in predators like siphonophores, armed with modular structures used exclusively for prey capture, this assumption becomes challenged. Our results show that siphonophores can evolve new prey type specializations and generalism by modifying the morphological states, modes of evolution, and genetic correlations among the parts of their prey capture apparatus. These findings demonstrate how studying open-ocean non-bilaterian predators can reveal novel patterns and mechanisms in the evolution of specialization. Understanding these evolutionary processes is fundamental to the study of food web structure and complexity.

@@

  Most animal predators use specific structures to capture and subdue prey. Raptors have claws and beaks, snakes have fangs, wasps have stingers, and cnidarians have nematocyst-laden tentacles. The functional morphology of these structures is critical to their ability to successfully capture prey [@schmitz2017predator]. Long-term adaptive evolution in response to the defense mechanisms of the prey (*e.g.*, avoidance, escape, protective barriers) leads to modifications that can counter those defenses. The more specialized the diet of a predator is, the more specialized its structures need to be to efficiently overcome the challenges posed by the prey. Characterizing the relationships between morphology and predatory specialization is necessary to understand how the phenotypic diversity of predators determines food web structure. However, for many clades of predators, there is scarce knowledge on how these specializations evolved. The primary questions we set out to answer are: how do predator specialists and generalists evolve, and how does predatory specialization shape morphological evolution?

  Siphonophores (Cnidaria: Hydrozoa) are a clade of predatory, gelatinous, colonial organisms that swim in the open ocean, feeding on a wide diversity of prey (often fish, crustaceans, and jellyfish). Siphonophores carry modular structures that are exclusively used for prey capture: the tentilla (Fig. \@ref(figure1)). The tentilla have great morphological variation across species [@mapstone2014global]. Together with their well understood function, this makes them an ideal system to study the relationships between functional traits and prey specialization. Like a head of coral, a siphonophore is a colony bearing many feeding polyps (Fig. \@ref(figure1)). Each feeding polyp has a single tentacle, which branches into a series of tentilla (side branches). Like other cnidarians, siphonophores capture prey with nematocysts, harpoon-like stinging capsules borne within specialized cells known as cnidocytes. Unlike the prey capture apparatus of most other cnidarians, siphonophore tentacles carry their cnidocytes in extremely complex and organized batteries [@skaer1988formation] which are located in their tentilla. While nematocyst batteries and clusters in other cnidarians are simple static scaffolds for cnidocytes, siphonophore tentilla have their own reaction mechanism, triggered upon encounter with prey. When it fires, a tentillum undergoes an extremely fast conformational change that wraps it around the prey, maximizing the surface area of contact for nematocysts to fire on the prey [@Mackie:1987uy]. In addition, some species have elaborate fluorescent and bioluminescent lures on their tentilla to attract prey with aggressive mimicry [@purcell1980influence;@haddock2005bioluminescent;@haddock2015fluorescent].

@@

*Novel phylogenetic relationships* –  In order to analyze the relationships between morphology and diet across the evolutionary history of siphonophores, we needed a siphonophore phylogeny that had broader taxonomic sampling than was available in previously published analyses. We first inferred a new tree with the needed taxon sampling with publicly available ribosomal RNA genes (18S & 16S) and new data from one species. This tree is essentially an extended version of that published in [@dunn2005molecular], and the two are congruent. We then compared the new extended ribosomal RNA tree to a recently published siphonophore transcriptome phylogeny [@munro2018improved]. The topology of the extended ribosomal RNA tree recapitulates the resolved nodes in [@dunn2005molecular] and most of the nodes in [@munro2018improved]. Only 5 nodes in the unconstrained tree inference were incongruent with the [@munro2018improved] transcriptome tree, 4 of them were poorly supported (bootstrap values <84), and only one of them was strongly supported (*Frillagalma vityazi*-*Nanomia bijuga*, 100 bootstrap support). We constrained the incongruent nodes to the [@munro2018improved] topology during estimation of the constrained 18S+16S tree inference (Fig. \@ref(figure4)). Since the transcriptome-based placement of *Nanomia bijuga* is more consistent with the morphological data, that relationship was also constrained. Moreover, with the inclusion of sequences from *Stephanomia amphytridis* and multiple *Erenna* species, our tree reveals a novel sister relationship between the genus *Erenna* and *Stephanomia*.

![\label{figure4} Bayesian time-tree inferred from 18S + 16S concatenated sequences and constrained to be congruent with a published transcriptome phylogeny. Branch lengths estimated using a relaxed molecular clock. Species names in red indicate replicated representation in the morphology data. All data was publicly available, apart from new sequences produced for Thermopalia taraxaca and Frillaglma vityazi. Nodes labeled with Bayesian posteriors (BP). Green circles indicate BP = 1. Blue circles indicate nodes constrained to be congruent with Munro *et al.* (2018). Tips with black squares indicate the species with transcriptomes used in Munro *et al.* (2018). Tips with purple squares indicate genus-level correspondence to taxa included in Munro *et al.* (2018). The main clades are labeled: in black for described taxonomic units, and in grey for operational phylogenetic designations.](Figures_final/figure4.png){height=600px}

@@

*Evolutionary associations between diet and tentillum morphology* – We reconstructed the evolutionary history of feeding guilds using stochastic mapping on the new phylogeny. Our reconstructions do not recover generalism as the ancestral siphonophore diet. None of the transitions in diet are consistent with scenario 1 (specialists evolving from generalists). Feeding guild specializations have shifted from an alternative ancestral state at least five times, consistent with instances supporting scenario 2 (specialists evolving to feed on a different resource). We also recover multiple independent origins of generalism from specialist ancestors (Fig. \@ref(figure5)). Large crustacean specialists evolve into generalists twice independently, consistent with instances of scenario 3 (generalists evolving from specialists). This finding is particularly compelling given in that it is the opposite of known biases in ancestral state reconstruction. [@nosil2005testing] found that such methods tend to infer higher transition rates toward the more frequent state. In this case, that would lead to a bias for an increased rate of transition from generalists (the rarer state across the tips) to specialists (the more common state across the tips). We observe the opposite, indicating strong evidence that these generalists are indeed a derived state.

To test whether measured morphological characters evolved in association with shifts in feeding ecology, we analyzed the evolutionary history of each character on the phylogeny, with the feeding guilds reconstructed on it as hypothetical selective regimes. We fit and compared alternative evolutionary models for each continuous character. The models compared were the Brownian Motion (BM) model of neutral divergent evolution [@martins1996phylogenies], the Ornstein-Uhlenbeck (OU) model of stabilizing selection around a single fitted optimum state [@uhlenbeck1930theory;@butler2004phylogenetic], and an OU model with multiple optima corresponding to each reconstructed selective regime (feeding guild). The model comparison shows that out of 30 characters, 10 show significantly stronger support for the diet-driven multi-optima OU model (S15). These characters include terminal filament nematocyst size and shape, involucrum length, elastic strand width, and heteroneme number. Most of these characters are found exclusively in Tendiculophora, thus this may reflect processes that could be unique to this subtree. Five characters including cnidoband length, cnidoband shape, and haploneme length show maximal support for a diet-driven single-optimum OU model. The remaining 15 characters support BM (or OU with marginal AICc difference with BM).

@@

   In addition to studying correlations with prey type presence/absence, we also tested for correlations between morphological characters and shifts in prey selectivity using phylogenetic linear models. Prey selectivity values were calculated from [@purcell1981dietary] by contrasting the gut content frequencies to the corresponding environmental abundances of prey. We found that fish selectivity is associated with increased number of heteronemes per tentillum, increased roundness of nematocysts (desmonemes and haplonemes), larger heteronemes, reduced heteroneme/cnidoband length ratios, smaller rhopalonemes, lower haploneme surface area to volume ratio (SA/V), and larger the cnidoband, elastic strand, pedicle and tentacle widths. Decapod-selective diets were associated with increasing cnidoband size and coiledness, haploneme row number, elastic strand width, and heteroneme number. Copepod-selective diets evolved in association with smaller heteroneme and total nematocyst volumes, smaller cnidobands, rounder rhopalonemes, elongated heteronemes, narrower haplonemes with higher SA/V ratios, and smaller heteronemes, tentacles, pedicles, and elastic strands. Selectivity for ostracods was associated with reductions in size and number of heteroneme nematocysts, cnidoband size, number of haploneme rows, heteroneme numbers, and cnidoband coiledness. Heteroneme length and elongation also correlated negatively with chaetognath selectivity (S21). These results indicate that not only diet but also differential feeding selectivity has evolved in correlation with changes in the prey capture apparatus of siphonophores.

@@

*Evolution of relationships between characters with diet* – Phenotypic integration results in correlation patterns between morphological characters and their rates of evolution. To study these patterns, we fit a set of evolutionary variance-covariance matrices [@revell2009phylogenetic]. The quantitative characters we measured from tentilla and their nematocysts are highly correlated. The results indicate that the dimensionality (number of independent axes of variation) of tentillum morphology is low, that many traits are associated with size, but that nematocyst arrangement and shape are independent of it (S4). The variance-covariance matrices (S36-38) are congruent with the abundant positive correlations observed among simple measurement characters in S3. This analysis more clearly reveals the diagonal blocks that constitute the evolutionary modules, such as the heteroneme block, the terminal filament nematocyst block, and the cnidoband-pedicle-tentacle block. These results were not sensitive to the transformation of inapplicable states and taxon sampling. These results indicate that siphonophore tentilla and nematocysts are phenotypically integrated and co-evolve within discrete evolutionary modules.

 In order to test whether rate covariance matrices changed with evolutionary shifts in feeding guild regimes, we compared the rate covariance terms between characters across the subtrees occupied by the different feeding guild regimes (S41). We found that half (48%) of the character pairs presented significantly distinct correlation coefficients across different regimes (S39), indicating that the mode of phenotypic integration also shifts with trophic niche. When contrasting the regime-specific rate correlation matrices to the whole-tree matrix, we were able to identify the character dependencies that are unique to each predatory niche (S42). These results indicate that the evolutionary dependencies in these integrated modules are changing across the phylogeny, and evolving together with changes in prey type specializations.

 We were able to identify specific character correlations that shifted with the evolution of new diets. Under the majority of stochastic character mapping outcomes, large crustacean specialists are the ancestral feeding regime, and all other feeding regimes evolve from this ancestral specialization. Compared to the rate correlation matrix estimated over the whole tree, large crustacean specialists present strong negative correlations between haploneme elongation and heteroneme size, and between rhopaloneme elongation and tentillum size, as well as with involucrum length. Within generalist clades (*Forskalia* and the *Agalma-Athorybia* clade), terminal filament nematocyst (desmonemes and rhopalonemes) sizes became negatively correlated with the sizes of most characters, meaning that as some tentilla became larger, their individual terminal nematocysts became smaller, observed to the extreme in *Agalma*. In addition, heteroneme and rhopaloneme elongation became positively correlated with cnidoband size. When large crustacean specialists switched to small crustacean prey in *Cordagalma* and calycophorans, haploneme size became inversely correlated with heteroneme elongation, which in turn developed a strong positive relationship with tentillum size. In other words, as tentilla get smaller in this group, heteronemes get shorter and haplonemes get larger. The extremes of this gradient can be seen in *Cordagalma* and *Hippopodius*. With the evolution of fish prey specialization in cystonects and within Clade B (Fig. \@ref(figure1)), haploneme elongation became negatively correlated with heteroneme elongation (signal driven by Clade B, since cystonects lack tentacular heteronemes), and the surface area to volume ratio of haploneme nematocysts switched from a strong negative relationship with cnidoband size (found in every other regime) to a positive correlation. Gelatinous specialization, albeit appearing only once in our tree, also carries a unique signature in character rate correlation shifts, with an increase in the strength of the correlation between heteroneme shape and shaft width, consistent with the appearance of birrhopaloid nematocysts with swollen shafts that are likely effective at anchoring gelatinous tissue (see reference to Narcomedusae nematocysts in [@purcell1988correlation]).

@@

Several studies [@futuyma1988evolution;@siddall1993phylogeny;@hougen1994interactions; @robinson1996trade;@kelley1998specialization] have suggested that resource specialization is an irreversible dead-end due to the constraints posed by extreme phenotypic specialization. Our results show that this is not the case for siphonophores, where the prey type on which they specialize has shifted at least 5 times. We find no support for any transitions from generalist to specialist (scenario 1, as described in the Introduction). We do find support for at least 3 instances of specialists switching from one prey type to another prey type, (scenario 2) and two switches from specialist to generalist (scenario 3). This is consistent with the findings of recent studies on phytophagous insects [@nosil2002transition], where the rate of evolution from generalists to specialists is comparable to the reverse, thus specialization does not limit further evolution. Our results are also consistent with analyses of lepidopterans [@hardy2014specialization], where specialized resource switching is the primary transition type while niche breadth remains fairly constant. The evolutionary history of tentilla shows that siphonophores are an example of trophic niche diversification via morphological innovation and evolution, which allowed transitions between specialized trophic niches. In more familiar predators, the prey capture apparatus is well integrated in the body (such as claws and jaws), leading to trade-offs and whole body adaptations to feeding specialization. The extreme modularity of the siphonophore prey capture apparatus could release them from the constraints typically imposed by adaptation to ecological specialization. This evolutionary mechanism is particularly important in a deep open ocean ecosystem, which is a relatively homogeneous physical environment, where the primary niche heterogeneity available is the potential interactions between organisms [@robison2004deep].

@@

Most studies on the evolution of predation have focused on vertebrate systems with a prey capture apparatus (such as jaws and limbs) deeply integrated into their bodies, serving multiple functions. This has led to a biased understanding of the evolutionary outcomes of specialization, where extreme morphological evolution constrains further shifts in their ecology. Siphonophores differ in many ways from commonly-known predators, using modular weapons for prey capture (the tentilla) that are fully decoupled from other structures and body functions. Our analysis of the evolutionary history of dietary specialization and morphological change in these elusive animals has revealed notable deviations from traditional expectations. While much of the feeding ecology literature focuses on how predatory generalists evolve into predatory specialists, in siphonophores we find predatory specialists can evolve into generalists, and that specialists on one prey type have directly evolved into specialists on other prey types. We find that the character states, evolutionary optima, and genetic correlations of many morphological characters have evolved following these ecological shifts. Our extended morphological characterization shows that the relationships between form and ecology hold across a large set of siphonophore taxa and characters. These findings are central to understanding the evolutionary mechanisms driving the emergence of food web complexity.

@@

*Tentillum morphology* – The morphological work was carried out on siphonophore specimens fixed in 4% formalin from the Yale Peabody Museum Invertebrate Zoology (YPM-IZ) collection (accession numbers in Dryad repository). These specimens were collected intact across many years of fieldwork expeditions, using blue-water diving [@haddock2005scientific], remotely operated vehicles (ROVs), plankton net trawls, and human-operated submersibles. Tentacles were dissected from non-larval gastrozooids, sequentially dehydrated into 100% ethanol, cleared in methyl salicylate, and mounted onto slides with Canada Balsam or Permount mounting media. The slides were imaged as tiled z-stacks using differential interference contrast (DIC) on an automated stage at YPM-IZ (with the assistance of Daniel Drew and Eric Lazo-Wasem) and with laser point confocal microscopy using a 488 nm Argon laser that excited autofluorescence in the tissues. Thirty characters (defined in S5) were measured using Fiji [@collins2007imagej;@schindelin2012fiji]. We did not measure the lengths of contractile structures (terminal filaments, pedicles, gastrozooids, and tentacles) since they are too variable to quantify. We measured at least one specimen for 96 different species (raw data available in Dryad). Of these, we selected 38 focal species across clades based on specimen availability and phylogenetic representation. Three to five tentacle specimens from each one of these selected species were measured to capture intraspecific variation.

*Siphonophore phylogeny* – While the main goal of this work is not to elucidate a novel phylogeny for Siphonophora, we did expand on the most recent transcriptome based phylogeny [@munro2018improved] to accommodate a larger taxon sampling. In order to do this, we ran a constrained analysis on an extensive 18S+16S dataset. The phylogenetic analysis included 55 siphonophore species and 6 outgroup cnidarian species (*Clytia hemisphaerica*, *Hydra circumcincta*, *Ectopleura dumortieri*, *Porpita porpita*, *Velella velella*, *Staurocladia wellingtoni*). The gene sequences we used in this study are available online (accession numbers in Dryad repository). Some of the sequences we used were accessioned in [@dunn2005molecular], and others we extracted from the transcriptomes in [@munro2018improved]. Two new 16S sequences for *Frillagalma vityazi* (MK958598) and *Thermopalia* sp. (MK958599) sequenced by Lynne Christianson using the primers from [@cunningham1993molecular] (read 3' to 5' F: TCGACTGTTTACCAAAAACATAGC , R:  ACGGAATGAACTCAAATCATGTAAG) were included and accessioned to NCBI. Additional details on the phylogenetic inference methods are available in the Supplementary Methods.

Given the broader sequence sampling of the transcriptome phylogeny, we ran constrained inferences (using both ML and Bayesian approaches, which produced fully congruent topologies (S8 and S10)) after fixing the 5 nodes that were incongruent with the topology of the consensus tree in [@munro2018improved]. This topology was then used to inform a Bayesian relaxed molecular clock time-tree in RevBayes, using a birth-death process (sampling probability calculated from the known number of described siphonophore species) to generate ultrametric branch lengths (S11-12). Scripts available in the Dryad repository.

@@

  We also extracted copepod prey length data from [@purcell1984functions]. To calculate specific prey selectivities, we extracted quantitative diet and zooplankton composition data from [@purcell1981dietary], matched each diet assessment to each prey field quantification by site, calculated Ivlev’s electivity indices [@jacobs1974quantitative], and averaged those by species (data available in the Dryad repository).

>>>>

# Classification: Biological Sciences {-}

@@

## ORCID: Alejandro Damian-Serrano: 0000-0002-2544-0489; Steven H.D. Haddock: 0000-0001-9420-4482; Casey Dunn: 0000-0003-0628-5150

\‡ Corresponding author: Alejandro Damian-Serrano. Address: 165 Prospect St., New Haven, CT, 06511. Phone: (401) 441-9613. Email: alejandro.damianserrano@yale.edu.

Siphonophores, nematocysts, predation, specialization, character evolution

\newpage

@@

@@

Predatory specialization is often associated with the evolution of modifications in the morphology of the prey capture apparatus. Specialization has been considered an evolutionary ‘dead-end’ due to the constraints associated with these morphological changes. However, in predators like siphonophores, armed with modular structures used exclusively for prey capture, this assumption is challenged. Our results show that siphonophores can evolve generalism and new prey-type specializations by modifying the morphological states, modes of evolution, and genetic correlations between the parts of their prey capture apparatus. These findings demonstrate how studying open-ocean non-bilaterian predators can reveal novel patterns and mechanisms in the evolution of specialization. Understanding these evolutionary processes is fundamental to the study of food-web structure and complexity.

@@

  Most animal predators use specific structures to capture and subdue prey. Raptors have claws and beaks, snakes have fangs, wasps have stingers, and cnidarians have nematocyst-laden tentacles. The functional morphology of these structures is critical to their ability to successfully capture prey [@schmitz2017predator]. Long-term adaptive evolution in response to the defense mechanisms of the prey (*e.g.*, avoidance, escape, protective barriers) leads to modifications that can counter those defenses. The more specialized the diet of a predator is, the more specialized its structures need to be to efficiently overcome the challenges posed by the prey. Characterizing the relationships between morphology and predatory specialization is necessary to understand how the phenotypic diversity of predators determines food-web structure. However, for many clades of predators, there is scarce knowledge on how these specializations evolved. The primary questions we set out to answer are: how do predator specialists and generalists evolve, and how does predatory specialization shape morphological evolution?

  Siphonophores (Cnidaria: Hydrozoa) are a clade of gelatinous, colonial organisms that swim in the open ocean, feeding on a wide diversity of prey (often fish, crustaceans, and jellyfish). Siphonophores carry modular structures that are exclusively used for prey capture: the tentilla (Fig. \@ref(figure1)). The tentilla have great morphological variation across species [@mapstone2014global]. Together with their well understood function, this makes them an ideal system to study the relationships between functional traits and prey specialization. Like a head of coral, a siphonophore is a colony bearing many feeding polyps (Fig. \@ref(figure1)). Each feeding polyp has a single tentacle, which branches into a series of tentilla (side branches). Like other cnidarians, siphonophores capture prey with nematocysts, harpoon-like stinging capsules borne within specialized cells known as cnidocytes. Unlike the prey capture apparatus of most other cnidarians, siphonophore tentacles carry their cnidocytes in extremely complex and organized batteries [@skaer1988formation] which are located in their tentilla. While nematocyst batteries and clusters in other cnidarians are simple static scaffolds for cnidocytes, siphonophore tentilla have their own reaction mechanism, triggered upon encounter with prey. When it fires, a tentillum undergoes an extremely fast conformational change that wraps it around the prey, maximizing the surface area of contact for nematocysts to fire on the prey [@Mackie:1987uy]. In addition, some species have elaborate fluorescent and bioluminescent lures on their tentilla to attract prey with aggressive mimicry [@purcell1980influence;@haddock2005bioluminescent;@haddock2015fluorescent].

@@

*Novel phylogenetic relationships* –  In order to analyze the relationships between morphology and diet across the evolutionary history of siphonophores, we generated a siphonophore phylogeny that had broader taxonomic sampling than was available in previously published analyses. We first inferred a new tree with the needed taxon sampling with publicly available ribosomal RNA genes (18S & 16S) and new data from one species. This tree is essentially an extended version of that published in [@dunn2005molecular], and the two are congruent. We then compared the new extended ribosomal RNA tree to a recently published siphonophore transcriptome phylogeny [@munro2018improved]. The topology of the extended ribosomal RNA tree recapitulates the resolved nodes in [@dunn2005molecular] and most of the nodes in [@munro2018improved]. Only five nodes in the unconstrained tree inference were incongruent with the [@munro2018improved] transcriptome tree, with four of them poorly supported (bootstrap values <84), and only one of them strongly supported (*Frillagalma vityazi*-*Nanomia bijuga*, 100 bootstrap support). We constrained the incongruent nodes to the [@munro2018improved] topology during estimation of the constrained 18S+16S tree inference (Fig. \@ref(figure4)). Since the transcriptome-based placement of *Nanomia bijuga* is more consistent with the morphological data, that relationship was also constrained. Moreover, with the inclusion of sequences from *Stephanomia amphytridis* and multiple *Erenna* species, our tree reveals a novel sister relationship between the genus *Erenna* and *Stephanomia*.

![\label{figure4} Bayesian time-tree inferred from 18S + 16S concatenated sequences and constrained to be congruent with a published transcriptome phylogeny. Branch lengths estimated using a relaxed molecular clock. Species names in red indicate replicated representation in the morphology data. All data were publicly available, apart from new sequences produced for Thermopalia taraxaca and Frillaglma vityazi. Nodes labeled with Bayesian posteriors (BP). Green circles indicate BP = 1. Blue circles indicate nodes constrained to be congruent with Munro *et al.* (2018). Tips with black squares indicate the species with transcriptomes used in Munro *et al.* (2018). Tips with purple squares indicate genus-level correspondence to taxa included in Munro *et al.* (2018). The main clades are labeled: with black bars for described taxonomic units, and grey bars for operational phylogenetic designations.](Figures_final/figure4.png){height=600px}

@@

*Evolutionary associations between diet and tentillum morphology* – We reconstructed the evolutionary history of feeding guilds using stochastic mapping on the new phylogeny (Fig. \@ref(figure5)). Our reconstructions do not recover generalism as the ancestral siphonophore diet. None of the transitions in diet are consistent with scenario 1 (specialists evolving from generalists). Feeding guild specializations have shifted from an alternative ancestral state at least five times, consistent with instances supporting scenario 2 (specialists evolving to feed on a different resource). We also recover multiple independent origins of generalism from specialist ancestors (Fig. \@ref(figure5)). Large crustacean specialists evolve into generalists twice independently, consistent with instances of scenario 3 (generalists evolving from specialists). This finding is particularly compelling given in that it is the opposite of known biases in ancestral state reconstruction. [@nosil2005testing] found that such methods tend to infer higher transition rates toward the more frequent state. In this case, that would lead to a bias for an increased rate of transition from generalists (the rarer state across the tips) to specialists (the more common state across the tips). We observe the opposite, indicating strong evidence that these generalists are indeed a derived state.

To test whether measured morphological characters evolved in association with shifts in feeding ecology, we analyzed the evolutionary history of each character on the phylogeny, with the feeding guilds reconstructed on it as hypothetical selective regimes. We fit and compared alternative evolutionary models for each continuous character. The models compared were the Brownian Motion (BM) model of neutral divergent evolution [@martins1996phylogenies], the Ornstein-Uhlenbeck (OU) model of stabilizing selection around a single fitted optimum state [@uhlenbeck1930theory;@butler2004phylogenetic], and an OU model with multiple optima (OUm) corresponding to each reconstructed selective regime (feeding guild). The model comparison shows that out of 30 characters, 10 show significantly stronger support for the diet-driven multi-optima OU model (S15). These characters include terminal filament nematocyst size and shape, involucrum length, elastic strand width, and heteroneme number. Most of these characters are found exclusively in Tendiculophora, thus this may reflect processes that could be unique to this subtree. Five characters including cnidoband length, cnidoband shape, and haploneme length show maximal support for a diet-driven single-optimum OU model. The remaining 15 characters support BM (or OU with marginal AICc difference with BM).

@@

   In addition to studying correlations with prey type presence/absence in the diet, we also tested for correlations between morphological characters and shifts in prey selectivity using phylogenetic linear models. Prey selectivity values were calculated from [@purcell1981dietary] by contrasting the gut content frequencies to the corresponding environmental abundances of prey. We found that fish selectivity is associated with increased number of heteronemes per tentillum, increased roundness of nematocysts (desmonemes and haplonemes), larger heteronemes, reduced heteroneme/cnidoband length ratios, smaller rhopalonemes, lower haploneme surface area to volume ratio (SA/V), and larger the cnidoband, elastic strand, pedicle and tentacle widths. Decapod-selective diets were associated with increasing cnidoband size and coiledness, haploneme row number, elastic strand width, and heteroneme number. Copepod-selective diets evolved in association with smaller heteroneme and total nematocyst volumes, smaller cnidobands, rounder rhopalonemes, elongated heteronemes, narrower haplonemes with higher SA/V ratios, and smaller heteronemes, tentacles, pedicles, and elastic strands. Selectivity for ostracods was associated with reductions in size and number of heteroneme nematocysts, cnidoband size, number of haploneme rows, heteroneme numbers, and cnidoband coiledness. Heteroneme length and elongation also correlated negatively with chaetognath selectivity (S21). These results indicate that not only diet but also differential feeding selectivity has evolved in correlation with changes in the prey capture apparatus of siphonophores.

@@

*Evolution of relationships between characters with diet* – Phenotypic integration results in correlation patterns between morphological characters and their rates of evolution. To study these patterns, we fit a set of evolutionary variance-covariance matrices [@revell2009phylogenetic]. The quantitative characters we measured from tentilla and their nematocysts are highly correlated. The results indicate that the dimensionality (number of independent axes of variation) of tentillum morphology is low, that many traits are associated with size, but that nematocyst arrangement and shape are independent of it (S4). The variance-covariance matrices (S21-23) are congruent with the abundant positive correlations observed among simple measurement characters in S3. This analysis more clearly reveals the diagonal blocks that constitute the evolutionary modules, such as the heteroneme block, the terminal filament nematocyst block, and the cnidoband-pedicle-tentacle block. These results were not sensitive to the transformation of inapplicable states and taxon sampling. These results indicate that siphonophore tentilla and nematocysts are phenotypically integrated and co-evolve within discrete evolutionary modules.

  In order to test whether rate covariance matrices changed with evolutionary shifts in feeding guild regimes, we compared the rate covariance terms between characters across the subtrees occupied by the different feeding guild regimes (S21). We found that half (48%) of the character pairs presented significantly distinct correlation coefficients across different regimes (S19), indicating that the mode of phenotypic integration also shifts with trophic niche. When contrasting the regime-specific rate correlation matrices to the whole-tree matrix and to the preceding ancestral regime matrix, we were able to identify the character dependencies that are unique to each predatory niche (S22-23).

 We were able to identify specific character correlations that shifted with the evolution of new diets. Under the majority of stochastic character mapping outcomes, large crustacean specialists are the ancestral feeding regime, and all other feeding regimes evolve from this ancestral specialization. Compared to the rate correlation matrix estimated over the whole tree, large crustacean specialists present strong negative correlations between haploneme elongation and heteroneme size, and between rhopaloneme elongation and tentillum size, as well as with involucrum length. Within generalist clades (*Forskalia* and the *Agalma-Athorybia* clade), terminal filament nematocyst (desmoneme and rhopaloneme) sizes became negatively correlated with the sizes of most characters, meaning that as some tentilla became larger, their individual terminal nematocysts became smaller, observed to the extreme in *Agalma*. In addition, heteroneme and rhopaloneme elongation became positively correlated with cnidoband size. When large crustacean specialists switched to small crustacean prey in *Cordagalma* and calycophorans, haploneme size became inversely correlated with heteroneme elongation, which in turn developed a strong positive relationship with tentillum size. In other words, as tentilla get smaller in this group, heteronemes get shorter and haplonemes get larger. The extremes of this gradient can be seen in *Cordagalma* and *Hippopodius*. With the evolution of fish prey specialization in cystonects and within Clade B (Fig. \@ref(figure1)), haploneme elongation became negatively correlated with heteroneme elongation (signal driven by Clade B, since cystonects lack tentacular heteronemes), and the surface area to volume ratio of haploneme nematocysts switched from a strong negative relationship with cnidoband size (found in every other regime) to a positive correlation. This is consistent with Clade B haplonemes becoming rounder, more similar to Cystonect haplonemes specialized in fish prey penetration and envenomation. Gelatinous specialization, albeit appearing only once in our tree, also carries a unique signature in character rate correlation shifts, with an increase in the strength of the correlation between heteroneme shape and shaft width, consistent with the appearance of birrhopaloid nematocysts with swollen shafts that are likely effective at anchoring gelatinous tissue (see reference to Narcomedusae nematocysts in [@purcell1988correlation]).

@@

Several studies [@futuyma1988evolution;@siddall1993phylogeny;@hougen1994interactions; @robinson1996trade;@kelley1998specialization] have suggested that resource specialization can be an irreversible dead-end due to the constraints posed by extreme phenotypic specialization. Our results show that this is not the case for siphonophores, where the prey type on which they specialize has shifted at least 5 times. We find no support for any transitions from generalist to specialist (scenario 1, as described in the Introduction). We do find support for at least 3 instances of specialists switching from one prey type to another prey type, (scenario 2) and two switches from specialist to generalist (scenario 3). This is consistent with the findings of recent studies on phytophagous insects [@nosil2002transition], where the rate of evolution from generalists to specialists is comparable to the reverse, thus specialization does not limit further evolution. Our results are also consistent with analyses of lepidopterans [@hardy2014specialization], where specialized resource switching is the primary transition type while niche breadth remains fairly constant. The evolutionary history of tentilla shows that siphonophores are an example of trophic niche diversification via morphological innovation and evolution, which allowed transitions between specialized trophic niches. In more familiar predators, the prey capture apparatus (such as claws and jaws) is well integrated in the body, leading to trade-offs and whole-body adaptations for feeding specialization. The extreme modularity of the siphonophore prey capture apparatus could release them from constraints typically imposed by adaptation to ecological specialization. This evolutionary mechanism is particularly important in a deep open ocean ecosystem, which is a relatively homogeneous physical environment, where the primary niche heterogeneity available is the potential interactions between organisms [@robison2004deep].

@@

Most studies on the evolution of predation have focused on vertebrate systems with an integrated feeding apparatus serving multiple functions. This has led to a biased understanding of the evolutionary outcomes of specialization, where extreme morphological evolution constrains further shifts in their ecology. Siphonophores differ in many ways from commonly-known predators, using modular weapons for prey capture (the tentilla) that are fully decoupled from other structures and body functions. Our analysis of the evolutionary history of dietary specialization and morphological change in these elusive animals has revealed notable deviations from traditional expectations. While much of the feeding ecology literature focuses on how predatory generalists evolve into predatory specialists, in siphonophores we find predatory specialists can evolve into generalists, and that specialists on one prey type have directly evolved into specialists on other prey types. We find that the character states, evolutionary optima, and genetic correlations of many morphological characters have evolved following these ecological shifts. We find that the relationships between form and ecology hold across a large set of siphonophore taxa and characters. These findings are central to understanding the evolutionary mechanisms driving the emergence of food web complexity.

@@

*Tentillum morphology* – The morphological work was carried out on siphonophore specimens fixed in 4% formalin from the Yale Peabody Museum Invertebrate Zoology (YPM-IZ) collection (accession numbers in Dryad repository). These specimens were collected intact across many years of fieldwork expeditions, using blue-water diving [@haddock2005scientific], remotely operated vehicles (ROVs), plankton net trawls, and human-operated submersibles. Tentacles were dissected from non-larval gastrozooids, sequentially dehydrated into 100% ethanol, cleared in methyl salicylate, and mounted onto slides with Canada Balsam or Permount mounting media. The slides were imaged as tiled z-stacks using differential interference contrast (DIC) on an automated stage at YPM-IZ (with the assistance of Daniel Drew and Eric Lazo-Wasem) and with laser point confocal microscopy using a 488 nm Argon laser that excited autofluorescence in the tissues. Thirty characters (defined in S1) were measured using Fiji [@collins2007imagej;@schindelin2012fiji]. We did not measure the lengths of contractile structures (terminal filaments, pedicles, gastrozooids, and tentacles) since they are too variable to quantify. We measured at least one specimen for 96 different species (raw data available in Dryad). Of these, we selected 38 focal species across clades based on specimen availability and phylogenetic representation. Three to five tentacle specimens from each one of these selected species were measured to capture intraspecific variation.

*Siphonophore phylogeny* – While the main goal of this work is not to elucidate a novel phylogeny for Siphonophora, we did expand on the most recent transcriptome based phylogeny [@munro2018improved] to accommodate a larger taxon sampling. In order to do this, we ran a constrained analysis on an extensive 18S+16S dataset. The phylogenetic analysis included 55 siphonophore species and 6 outgroup cnidarian species (*Clytia hemisphaerica*, *Hydra circumcincta*, *Ectopleura dumortieri*, *Porpita porpita*, *Velella velella*, *Staurocladia wellingtoni*). The gene sequences we used in this study are available online (accession numbers in Dryad repository). Some of the sequences we used were accessioned in [@dunn2005molecular], and others we extracted from the transcriptomes in [@munro2018improved]. Two new 16S sequences for *Frillagalma vityazi* (MK958598) and *Thermopalia* sp. (MK958599) sequenced by Lynne Christianson using the primers from [@cunningham1993molecular] (read 3' to 5' F: TCGACTGTTTACCAAAAACATAGC , R:  ACGGAATGAACTCAAATCATGTAAG) were included and accessioned to NCBI. Additional details on the phylogenetic inference methods can be found in the Supplementary Methods.

Unconstrained ML and Bayesian phylogenies were congruent (S2,S5). Given the broader sequence sampling of the transcriptome phylogeny, we ran constrained inferences (using both ML and Bayesian approaches, which produced fully congruent topologies (S4,S6)) after clamping the 5 nodes (S3) that were incongruent with the topology of the consensus tree in [@munro2018improved]. This topology was then used to inform a Bayesian relaxed molecular clock time-tree in RevBayes, using a birth-death process (sampling probability calculated from the known number of described siphonophore species) to generate ultrametric branch lengths (S7-8). Scripts and tree files available in the Dryad repository.

@@

We also extracted copepod prey length data from [@purcell1984functions]. To calculate specific prey selectivities, we extracted quantitative diet and zooplankton composition data from [@purcell1981dietary], matched each diet assessment to each prey field quantification by site, calculated Ivlev’s electivity indices [@jacobs1974quantitative], and averaged those by species (data available in Dryad repository).
